Surface-driven registration method for the structure-informed segmentation of diffusion MR images\tnoteref{t1}

### 2015-10-26 Publications/2015-NeuroImage/00-main.tex
fixes for version 1.2
updates before sending @zosso and @daducci
v1.3.0rc

% \documentclass[5p,authoryear,fleqn]{elsarticle}
\documentclass[3p,authoryear,fleqn]{elsarticle}
\journal{NeuroImage}
\usepackage[utf8]{inputenc}

% *** MISC UTILITY PACKAGES ***
\usepackage[mode=buildnew]{standalone}
\usepackage{gincltex}
\usepackage{glossaries}
\usepackage{booktabs} % to use \toprule
\usepackage{multirow}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage{xifthen}
\usepackage{xspace}
\usepackage{graphicx}
\graphicspath{{figures/}}

%\usepackage[cmex10,intlimits]{amsmath}
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage{algorithmic}
\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\argmax}{argmax}
\DeclareMathOperator{\argmin}{argmin}
\DeclareMathOperator{\diag}{diag}
\DeclareMathOperator{\dist}{dist}
\DeclareMathOperator{\const}{Const.}
\providecommand{\e}[1]{\ensuremath{\times 10^{#1}}}
\providecommand{\mdist}[2]{ \mathcal{D}_{#2}^2(\mathbf{#1}) }
\providecommand{\omegaset}{\ensuremath{\boldsymbol{\Omega}}}
\providecommand{\gammaset}{\ensuremath{\boldsymbol{\Gamma}}}
\providecommand{\regseg}{\emph{regseg}}
\providecommand{\Regseg}{\emph{Regseg}}
\providecommand{\lowb}{\textit{b0}}

@@

% *** PDF, URL AND HYPERLINK PACKAGES ***
%
\usepackage{url}
\usepackage[colorlinks=true,linkcolor=black, citecolor=blue, urlcolor=blue]{hyperref}
\usepackage{doi}

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor an-isot-ropy regis-tra-tion iso-tropic Free-Surfer %
align-ed pre-sent work-flow data-set data-sets}

@@

% http://tex.stackexchange.com/questions/56105/missing-references-heading
\def\bibsection{\section*{References}}
\begin{document}
\begin{frontmatter}

>>>>

%\documentclass[5p,authoryear,fleqn]{elsarticle}
\documentclass[3p,authoryear,fleqn]{elsarticle}
\journal{NeuroImage}
\usepackage[utf8]{inputenc}

\usepackage{xifthen}
\newboolean{review}
\setboolean{review}{true}

% *** MISC UTILITY PACKAGES ***
\usepackage[mode=buildnew]{standalone}
\usepackage{gincltex}
\usepackage{glossaries}
\usepackage{booktabs} % to use \toprule
\usepackage{multirow}

% *** PDF, URL AND HYPERLINK PACKAGES ***
%
\usepackage{url}
\usepackage[colorlinks=true]{hyperref}
% \usepackage[hyperref]{xcolor}
\usepackage{doi}

\usepackage{xspace}
\usepackage{graphicx}
\graphicspath{{figures/}}

%\usepackage[cmex10,intlimits]{amsmath}
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage{algorithmic}
\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\argmax}{argmax}
\DeclareMathOperator{\argmin}{argmin}
\DeclareMathOperator{\diag}{diag}
\DeclareMathOperator{\dist}{dist}
\DeclareMathOperator{\const}{Const.}
\providecommand{\e}[1]{\ensuremath{\cdot\text{10}^\text{#1}}}
\providecommand{\mdist}[2]{ \mathcal{D}_{#2}^2(\mathbf{#1}) }
\providecommand{\omegaset}{\ensuremath{\boldsymbol{\Omega}}}
\providecommand{\gammaset}{\ensuremath{\boldsymbol{\Gamma}}}
\providecommand{\regseg}{\emph{regseg}}
\providecommand{\Regseg}{\emph{Regseg}}
\providecommand{\lowb}{\textit{b0}}

@@

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor an-isot-ropy regis-tra-tion iso-tropic Free-Surfer %
align-ed pre-sent work-flow data-set data-sets micro-struc-tu-re}

@@

\ifthenelse{\boolean{review}}
{
  \usepackage[colorinlistoftodos,textsize=tiny, textwidth=1.2cm]{todonotes}
  \setlength{\marginparwidth}{1.5cm}
  \reversemarginpar
  % \setlength{\marginparwidth}{2.0cm}
  \newcommand{\revcomment}[2][]{\ifthenelse{\equal{#1}{}}{}{\todo{#1}}%
  \begingroup \color{blue!70} #2\endgroup}
}
{
  \newcommand{\todo}[2][]{}
  \newcommand{\revcomment}[2][]{#2}
}

% http://tex.stackexchange.com/questions/56105/missing-references-heading
\def\bibsection{\section*{References}}

\begin{document}
\hypersetup{linkcolor=black!60, citecolor=black!60, urlcolor=black!60}
\begin{frontmatter}

### 2015-10-26 Publications/2015-NeuroImage/02-abstract.tex
v1.3.0rc

% -*- root: 00-main.tex -*-
\newcomment[RV\#1(C.1)]{%
Current methods using diffusion MRI to map the microstructure and connectivity of the
  human brain increasingly require precise delineations of anatomical structures.
The problem is typically solved by projecting the structural information extracted
  in T1-weighted images, by means of nonlinear registration since diffusion
  MRI data present geometrical distortions.
Here we propose \regseg{}, which segments homogeneous regions within multivariate
  images by projecting a set of nested surfaces extracted from structural images.}
The surfaces evolve through a free-form deformation field supported by the B-spline basis
  to optimally map the contours onto the data in the target space.
We tested the functionality of \regseg{} using four digital phantoms warped with known and
  randomly generated deformations, where subvoxel accuracy was achieved.
\newcomment[RV\#2(C.11)]{%
We proposed a set of surfaces that define a segmentation model of the fractional anisotropy and
  the apparent diffusion coefficient maps derived from diffusion MRI datasets.}
These surfaces extracted from the same-subject T1-weighted image successfully segmented 16 real
  diffusion MRI datasets from the Human Connectome Project that were warped by realistic and nonlinear
  distortions.
We computed the misregistration error of the surfaces estimated by \regseg{} with respect
  to their theoretical location using the ground truth, thereby obtaining a 95\% CI of 0.56--0.66 mm
  distance between corresponding mesh vertices, which was below the 1.25 mm isotropic resolution of the images.
We also compared the performance of the proposed method with a widely used registration tool, which showed
  that \regseg{} outperformed this method in our settings.

>>>>

% -*- root: 00-main.tex -*-
\revcomment[RV\#1(C.1)\par RV\#2(C.10)]{%
Current methods using diffusion MRI to map the microstructure and connectivity of the
  human brain increasingly require precise delineations of anatomical structures.
The problem is typically solved by projecting the structural information extracted
  in T1-weighted images, by means of nonlinear registration since diffusion
  MRI data present geometrical distortions.
Here we propose \regseg{}, which segments homogeneous regions within multivariate
  images by projecting a set of nested surfaces extracted from structural images.}
The surfaces evolve through a free-form deformation field supported by the B-spline basis
  to optimally map the contours onto the data in the target space.
We tested the functionality of \regseg{} using four digital phantoms warped with known and
  randomly generated deformations, where subvoxel accuracy was achieved.
\revcomment[RV\#2(C.11)]{%
We proposed a set of surfaces that define a segmentation model of the fractional anisotropy and
  the apparent diffusion coefficient maps derived from diffusion MRI datasets.}
These surfaces extracted from the same-subject T1-weighted image successfully segmented 16 real
  diffusion MRI datasets from the Human Connectome Project that were warped by realistic and nonlinear
  distortions.
We computed the misregistration error of the surfaces estimated by \regseg{} with respect
  to their theoretical location using the ground truth, thereby obtaining a 95\% CI of 0.56--0.66 mm
  distance between corresponding mesh vertices, which was below the 1.25 mm isotropic resolution of the images.
We also compared the performance of the proposed method with a widely used registration tool, which showed
  that \regseg{} outperformed this method in our settings.

### 2015-10-26 Publications/2015-NeuroImage/03-introduction.tex
v1.3.0rc

% -*- root: 00-main.tex -*-
\section{Introduction}\label{sec:regseg-intro}
\Gls*{dmri} enables the mapping of microstructure \citep{basser_microstructural_1996}
  and connectivity \citep{craddock_imaging_2013} of the human brain \emph{in-vivo}.
It is generally acquired using \gls*{epi} schemes, since they are very fast at
  scanning a large sequence of images called \glspl*{dwi}.
Each \gls*{dwi} is sensitized with a gradient to probe proton diffusion in a certain
  \newcomment[RV\#2(C.13)]{orientation}.
Subsequent processing involves describing the local microstructure with one available
  model, which ranges from the early \gls*{dti} proposed by \cite{basser_microstructural_1996}
  to current models such as \citep{daducci_accelerated_2015}.
The microstructural map is then used to draw the preferential \newcomment[RV\#2(C.13)]{orientations} of diffusion
  across the brain using tractography \citep{mori_threedimensional_1999}.
Finally, a graph representing the corresponding structural network is built using
  the regions of a cortical parcellation as nodes and the streamlines of
  tractography as edges \citep{hagmann_mapping_2008}.
\newcomment[RV\#2(C.17)]{%
The methodologies to solve reconstruction, tractography and network building
  require the delineation of the anatomy in the \gls*{dmri} space.
Moreover, current trends on reconstruction \citep{jeurissen_multitissue_2014} and
  tractography \citep{smith_anatomicallyconstrained_2012} are increasingly using
  structural information to improve the microstructural mapping and fiber-tracking.}

\newcomment[RV\#2(C.15)]{%
Possibly, the earliest structural information incorporated to aid \gls*{dmri} processing
  is the \gls*{wm} mask used as a termination criteria for tractography.
The standardized procedure to obtain this mask was thresholding the \gls*{fa} map.
However, the mask and subsequent analyses are highly dependent on the threshold set
  \citep{taoka_fractional_2009}.
\todo{RV\#2(C.16)}
To overcome the unreliability of \gls*{fa} thresholding, and to broaden
  \gls*{wm} segmentation to brain tissue segmentation, a large number of
  methods have been proposed using \glspl*{dwi}, the \lowb{}, and \gls*{dti}-derived
  scalar maps such as \gls*{fa}, \gls*{adc} and others \citep{zhukov_level_2003,
  rousson_level_2004,jonasson_segmentation_2005,hadjiprocopis_unbiased_2005,liu_brain_2007,
  lu_segmentation_2008,han_experimental_2009}.}
\todo{RV\#2(C.14)}%
However, the precise segmentation of \gls*{dmri} is difficult for several reasons.
First, \gls{dmri} images have a resolution that is \newcomment[RV\#2(C.12)]{much lower}
  than that of the imaged microstructural features.
Therefore, voxels located in structural discontinuities are affected by partial
  voluming of the signal sources.
Second, the extremely low \gls*{snr} and the high dimensionality of the \glspl*{dwi} prevent
  their direct use in segmentation.
Third, the low contrast between \gls*{gm} and \gls*{wm} in the \lowb{} volumes also makes
  them unsuitable for brain tissue segmentation.

\newcomment[RV\#2(C.9)]{%
An alternative route to segmentation in native \gls*{dmri} space is the mapping of the
  structural information extracted from anatomical MR images, such as \gls*{t1}, using
  image registration techniques.
Generally, intra-subject registration of MR images of the brain involves only a linear
  mapping to compensate for head motion between scans.
However, \gls*{epi} introduces a geometrical distortion \citep{jezzard_correction_1995}
  that impedes the linear mapping from the structural space and warps the anatomy 
  identified on native \gls*{dmri} space \citep{irfanoglu_effects_2012}.
Numerous methods have been proposed to overcome this problem by incorporating information
  from extra MR acquisitions such as fieldmaps \citep{jezzard_correction_1995},
  \glspl*{dwi} with a different \gls*{pe} scheme \citep{cordes_geometric_2000,chiou_simple_2000},
  or \gls*{t2} images \citep{kybic_unwarping_2000,studholme_accurate_2000}.
These methods estimate the deformation field associated to \emph{\gls*{epi} distortions} and
  resample the \glspl*{dwi} onto a corrected \gls*{dmri} space.
The retrospective \gls*{epi} correction is an active field of research yielding frequent refinements 
  and combinations of the original methods, such as
  \citep{holland_efficient_2010,andersson_comprehensive_2012,
  irfanoglu_drbuddi_2015}. \todo{RV\#2(C.18)}
A standardized method to solve the remaining linear mapping between the corrected-\gls*{dmri}
  and the structural spaces is \emph{bbregister} \citep{greve_accurate_2009}.%
}

\newcomment[RV\#2(C.10)]{%
Here we present a segmentation and surface-to-volume registration method 
  named \regseg{}, and show its usefulness in incorporating the anatomical
  information into native \gls*{dmri} space to aid subsequent reconstruction,
  tractography and network building using a cortical parcellation.
The hypothesis underlying is that the registration and segmentation problems
  in \gls*{dmri} can be solved simultaneously.
To implement \regseg{} we first establish an active-contours without edges
  \citep{chan_active_2001} segmentation framework.
Several \emph{anatomically-correct} surfaces of the same subject initialize
  the 3D active-contours, which evolve searching for homogeneous regions in
  the multivariate target-image.
In the presented application to segment \gls*{dmri} data, we propose a bivariate 
  target-volume comprehending the \gls*{fa} and \gls*{adc} maps.
The evolution of the surfaces is supported by a B-spline basis, optimized
  iteratively using a descent approach driven by shape-gradients
  \citep{besson_dream2s_2003,herbulot_segmentation_2006}.
Therefore, \regseg{} establishes a registration framework that actually
  overcomes the nonlinear warping induced by \gls*{epi} distortions.%
}
\Regseg{} integrates the benefits of segmentation and registration methods together and
  exploits the multivariate nature of \gls*{dmri} data to contribute in the proposed
  application on three key aspects:
  1) it avoids the resampling the \glspl*{dwi} introduced by the correction of \gls*{epi}
    distortions, since processing can be performed on the native \gls*{dmri} space;
  2) the surfaces are typically extracted from the \gls*{t1} image of the same subject,
    therefore \regseg{} does not require additional MR acquisitions to the minimal
    \gls*{dmri} protocol; and
  3) \regseg{} increases the geometrical accuracy of the overall process.
In this paper, we first verify the functionality of the method and the \regseg{}
  implementation using a set of digital phantoms, demonstrating the sub-pixel accuracy
  in registration.
We then evaluate \regseg{} on real \gls*{dmri} datasets, using a derivation of our
  instrumentation framework \citep{esteban_simulationbased_2014} which simulates 
  known and realistic \gls*{epi} distortions.
We also compare \regseg{} and a nonlinear registration method to map the \lowb{} to 
  the corresponding \gls*{t2} image of the same subject.
This method is a derivation of the \gls*{t2b} correction method, reproducing the settings
  of an existing implementation included within a widely used diffusion processing software
  \citep{leemans_exploredti_2009}.
With this comparison we demonstrate how \regseg{} achieves higher accuracy with the
  simultaneous registration and segmentation process.

>>>>

% -*- root: 00-main.tex -*-
\section{Introduction}\label{sec:regseg-intro}
\Gls*{dmri} enables the mapping of microstructure \citep{basser_microstructural_1996}
  and connectivity \citep{craddock_imaging_2013} of the human brain \emph{in-vivo}.
It is generally acquired using \gls*{epi} schemes, since they are very fast at
  scanning a large sequence of images called \glspl*{dwi}.
Each \gls*{dwi} is sensitized with a gradient to probe proton diffusion in a certain
  \revcomment[RV\#2(C.13)]{orientation}.
Subsequent processing involves describing the local microstructure with one available
  model, which ranges from the early \gls*{dti} proposed by \cite{basser_microstructural_1996}
  to current models such as \citep{daducci_accelerated_2015}.
The microstructural map is then used to draw the preferential \revcomment[RV\#2(C.13)]{orientations} of diffusion
  across the brain using tractography \citep{mori_threedimensional_1999}.
Finally, a graph representing the corresponding structural network is built using
  the regions of a cortical parcellation as nodes and the streamlines of
  tractography as edges \citep{hagmann_mapping_2008}.
\revcomment[RV\#2(C.17)]{%
The methodologies to solve reconstruction, tractography and network building
  require the delineation of the anatomy in the \gls*{dmri} space.
Moreover, current trends on reconstruction \citep{jeurissen_multitissue_2014} and
  tractography \citep{smith_anatomicallyconstrained_2012} are increasingly using
  structural information to improve the microstructural mapping and fiber-tracking.}

\revcomment[RV\#2(C.15)]{%
Possibly, the earliest structural information incorporated to aid \gls*{dmri} processing
  is the \gls*{wm} mask used as a termination criteria for tractography.
The standardized procedure to obtain this mask was thresholding the \gls*{fa} map.
However, the mask and subsequent analyses are highly dependent on the threshold set
  \citep{taoka_fractional_2009}.
\todo{RV\#2(C.16)}
To overcome the unreliability of \gls*{fa} thresholding, and to broaden
  \gls*{wm} segmentation to brain tissue segmentation, a large number of
  methods have been proposed using \glspl*{dwi}, the \lowb{}, and \gls*{dti}-derived
  scalar maps such as \gls*{fa}, \gls*{adc} and others \citep{zhukov_level_2003,
  rousson_level_2004,jonasson_segmentation_2005,hadjiprocopis_unbiased_2005,liu_brain_2007,
  lu_segmentation_2008,han_experimental_2009}.}
\todo{RV\#2(C.14)}%
However, the precise segmentation of \gls*{dmri} is difficult for several reasons.
First, \gls{dmri} images have a resolution that is \revcomment[RV\#2(C.12)]{much lower}
  than that of the imaged microstructural features.
Therefore, voxels located in structural discontinuities are affected by partial
  voluming of the signal sources.
Second, the extremely low \gls*{snr} and the high dimensionality of the \glspl*{dwi} prevent
  their direct use in segmentation.
Third, the low contrast between \gls*{gm} and \gls*{wm} in the \lowb{} volumes also makes
  them unsuitable for brain tissue segmentation.

\revcomment[RV\#2(C.9)]{%
An alternative route to segmentation in native \gls*{dmri} space is the mapping of the
  structural information extracted from anatomical MR images, such as \gls*{t1}, using
  image registration techniques.
Generally, intra-subject registration of MR images of the brain involves only a linear
  mapping to compensate for head motion between scans.
However, \gls*{epi} introduces a geometrical distortion \citep{jezzard_correction_1995}
  that impedes the linear mapping from the structural space and warps the anatomy 
  identified on native \gls*{dmri} space \citep{irfanoglu_effects_2012}.
Numerous methods have been proposed to overcome this problem by incorporating information
  from extra MR acquisitions such as fieldmaps \citep{jezzard_correction_1995},
  \glspl*{dwi} with a different \gls*{pe} scheme \citep{cordes_geometric_2000,chiou_simple_2000},
  or \gls*{t2} images \citep{kybic_unwarping_2000,studholme_accurate_2000}.
These methods estimate the deformation field associated to \emph{\gls*{epi} distortions} and
  resample the \glspl*{dwi} onto a corrected \gls*{dmri} space.
The retrospective \gls*{epi} correction is an active field of research yielding frequent refinements 
  and combinations of the original methods, such as
  \citep{holland_efficient_2010,andersson_comprehensive_2012,
  irfanoglu_drbuddi_2015}. \todo{RV\#2(C.18)}
A standardized method to solve the remaining linear mapping between the corrected-\gls*{dmri}
  and the structural spaces is \emph{bbregister} \citep{greve_accurate_2009}.%
}

\revcomment[RV\#2(C.10)]{%
Here we present a segmentation and surface-to-volume registration method 
  named \regseg{}, and show its usefulness in incorporating the anatomical
  information into native \gls*{dmri} space to aid subsequent reconstruction,
  tractography and network building using a cortical parcellation.
The hypothesis underlying is that the registration and segmentation problems
  in \gls*{dmri} can be solved simultaneously.
To implement \regseg{} we first establish an active-contours without edges
  \citep{chan_active_2001} segmentation framework.
Several \emph{anatomically-correct} surfaces of the same subject initialize
  the 3D active-contours, which evolve searching for homogeneous regions in
  the multivariate target-image.
In the presented application to segment \gls*{dmri} data, we propose a bivariate 
  target-volume comprehending the \gls*{fa} and \gls*{adc} maps.
The evolution of the surfaces is supported by a B-spline basis, optimized
  iteratively using a descent approach driven by shape-gradients
  \citep{besson_dream2s_2003,herbulot_segmentation_2006}.
Therefore, \regseg{} establishes a registration framework that actually
  overcomes the nonlinear warping induced by \gls*{epi} distortions.%
}
\Regseg{} integrates the benefits of segmentation and registration methods together and
  exploits the multivariate nature of \gls*{dmri} data to contribute in the proposed
  application on three key aspects:
  1) it avoids the resampling the \glspl*{dwi} introduced by the correction of \gls*{epi}
    distortions, since processing can be performed on the native \gls*{dmri} space;
  2) the surfaces are typically extracted from the \gls*{t1} image of the same subject,
    therefore \regseg{} does not require additional MR acquisitions to the minimal
    \gls*{dmri} protocol; and
  3) \regseg{} increases the geometrical accuracy of the overall process.
In this paper, we first verify the functionality of the method and the \regseg{}
  implementation using a set of digital phantoms, demonstrating the sub-pixel accuracy
  in registration.
We then evaluate \regseg{} on real \gls*{dmri} datasets, using a derivation of our
  instrumentation framework \citep{esteban_simulationbased_2014} which simulates 
  known and realistic \gls*{epi} distortions.
We also compare \regseg{} and a nonlinear registration method to map the \lowb{} to 
  the corresponding \gls*{t2} image of the same subject.
This method is a derivation of the \gls*{t2b} correction method, reproducing the settings
  of an existing implementation included within a widely used diffusion processing software
  \citep{leemans_exploredti_2009}.
With this comparison we demonstrate how \regseg{} achieves higher accuracy with the
  simultaneous registration and segmentation process.
